---
title: "DIYtranscriptomics"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    source_code: embed
---

```{r setup, include=FALSE}
library(tidyverse)
library(reshape2)
library(tximport)
library(biomaRt)
library(ensembldb)
library(RColorBrewer) 
library(genefilter)
library(edgeR) 
library(matrixStats)
library(gplots) 
library(limma)
library(heatmaply)
library(DT)
library(gt)
library(plotly)
library(hrbrthemes)
library(EnsDb.Hsapiens.v86)
library(cowplot)
library(gprofiler2)
library(GSEABase)
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

Overview {data-icon="fa-signal"}
===================================== 

Row {data-height=400}
------------------------------------------------------------------------------

### Signal distribution: raw data

```{r raw data}
# importing data into R using Tximport package
library(tidyverse) # provides access to Hadley Wickham's collection of R packages for data science, which we will use throughout the course
library(tximport) # package for getting Kallisto results into R
library(ensembldb) #helps deal with ensembl
library(EnsDb.Hsapiens.v86) #replace with your organism-specific database package
targets <- read_tsv("studyDesign.txt")# read in your study design
path <- file.path(targets$sample, "abundance.h5") # set file paths to your mapped data
Tx <- transcripts(EnsDb.Hsapiens.v86, columns=c("tx_id", "gene_name"))
Tx <- as_tibble(Tx)
Tx <- dplyr::rename(Tx, target_id = tx_id)
Tx <- dplyr::select(Tx, "target_id", "gene_name")
Txi_gene <- tximport(path, 
                     type = "kallisto", 
                     tx2gene = Tx, 
                     txOut = FALSE, #determines whether your data represented at transcript or gene level
                     countsFromAbundance = "lengthScaledTPM",
                     ignoreTxVersion = TRUE)

library(reshape2) 
library(genefilter)
library(edgeR) 
library(matrixStats)

sampleLabels <- targets$sample
myDGEList <- DGEList(Txi_gene$counts)
log2.cpm <- cpm(myDGEList, log=TRUE)

log2.cpm.df <- as_tibble(log2.cpm, rownames = "geneID")
colnames(log2.cpm.df) <- c("geneID", sampleLabels)
log2.cpm.df.melt <- melt(log2.cpm.df)

p1 <- ggplot(log2.cpm.df.melt, aes(x=variable, y=value, fill=variable)) +
  geom_violin(trim = FALSE, show.legend = FALSE) +
  stat_summary(fun.y = "median", 
               geom = "point", 
               shape = 124, 
               size = 6, 
               color = "black", 
               show.legend = FALSE) +
  labs(y="log2 expression", x = "sample",
       title="Log2 Counts per Million (CPM)",
       subtitle="unfiltered, non-normalized",
       caption=paste0("produced on ", Sys.time()))

cpm <- cpm(myDGEList)
keepers <- rowSums(cpm>1)>=3 #user defined
myDGEList.filtered <- myDGEList[keepers,]

log2.cpm.filtered <- cpm(myDGEList.filtered, log=TRUE)
log2.cpm.filtered.df <- as_tibble(log2.cpm.filtered, rownames = "geneID") 
colnames(log2.cpm.filtered.df) <- c("geneID", sampleLabels)
log2.cpm.filtered.df.melt <- melt(log2.cpm.filtered.df)

p2 <- ggplot(log2.cpm.filtered.df.melt, aes(x=variable, y=value, fill=variable)) +
  geom_violin(trim = FALSE, show.legend = FALSE) +
  stat_summary(fun.y = "median", 
               geom = "point", 
               shape = 124, 
               size = 6, 
               color = "black", 
               show.legend = FALSE) +
  labs(y="log2 expression", x = "sample",
       title="Log2 Counts per Million (CPM)",
       subtitle="filtered, non-normalized",
       caption=paste0("produced on ", Sys.time()))

myDGEList.filtered.norm <- calcNormFactors(myDGEList.filtered, method = "TMM")
log2.cpm.filtered.norm <- cpm(myDGEList.filtered.norm, log=TRUE)
log2.cpm.filtered.norm.df <- as_tibble(log2.cpm.filtered.norm, rownames = "geneID")
colnames(log2.cpm.filtered.norm.df) <- c("geneID", sampleLabels)
log2.cpm.filtered.norm.df.melt <- melt(log2.cpm.filtered.norm.df)

p3 <- ggplot(log2.cpm.filtered.norm.df.melt, aes(x=variable, y=value, fill=variable)) +
  geom_violin(trim = FALSE, show.legend = FALSE) +
  stat_summary(fun.y = "median", 
               geom = "point", 
               shape = 124, 
               size = 6, 
               color = "black", 
               show.legend = FALSE) +
  labs(y="log2 expression", x = "sample",
       title="Log2 Counts per Million (CPM)",
       subtitle="filtered, TMM normalized",
       caption=paste0("produced on ", Sys.time()))

p1
```

### Signal distribution: filtered and normalized data

```{r processed data}
p3
```

### PCA plot

```{r PCA}
treatment <- factor(targets$treatment)

pca.res <- prcomp(t(log2.cpm.filtered.norm), scale.=F, retx=T)
pc.var<-pca.res$sdev^2
pc.per<-round(pc.var/sum(pc.var)*100, 1)

pca.res.df <- as_tibble(pca.res$x)
ggplot(pca.res.df, aes(x=PC1, y=PC2, color=targets$treatment)) +
  geom_point(size=4) +
  xlab(paste0("PC1 (",pc.per[1],"%",")")) + 
  ylab(paste0("PC2 (",pc.per[2],"%",")")) +
  labs(title="PCA plot",
       caption=paste0("produced on ", Sys.time())) +
  theme_bw()

```

Row {data-height=400}
------------------------------------------------------------------------------

### Filtered and normalized counts per million (CPM)

```{r filtered and normalized data}
library(tidyverse)
library(reshape2)
library(DT)
library(gt)
library(plotly)

mydata.df <- as_tibble(log2.cpm.filtered.norm, rownames = "geneSymbol")
colnames(mydata.df) <- c("geneSymbol", sampleLabels)

mydata.df <- mutate(mydata.df,
                    uninfected.AVG = (uninf_rep1 + uninf_rep2 + uninf_rep1)/3, 
                    crypto.wt.AVG = (crypto.wt_rep1 + crypto.wt_rep2 + crypto.wt_rep3)/3,
                    crypto.mut.AVG = (crypto.mut_rep1 + crypto.mut_rep2 + crypto.mut_rep3)/3,
                    #now make columns comparing each of the averages above that you're interested in
                    LogFC.crypto.wt_vs_uninfected = (crypto.wt.AVG - uninfected.AVG),
                    LogFC.crypto.mut_vs_uninfected = (crypto.mut.AVG - uninfected.AVG)) %>%
  mutate_if(is.numeric, round, 2)

datatable(mydata.df, 
          extensions = c('KeyTable', "FixedHeader"), 
          caption = 'Table 1: Filtered and normalized data',
          options = list(keys = TRUE, 
                         searchHighlight = TRUE, 
                         pageLength = 10, 
                         lengthMenu = c("10", "25", "50", "100")))
```

DEGs {data-orientation=columns data-icon="fa-random"}
===================================== 

Column {data-width=500}
------------------------------------------------------------------------------

### Volcano plot

```{r volcano plot}
library(tidyverse)
library(limma)
library(edgeR) 
library(gt)
library(DT) 
library(plotly) 

treatment <- factor(targets$treatment)
design <- model.matrix(~0 + treatment)
colnames(design) <- levels(treatment)
v.DEGList.filtered.norm <- voom(myDGEList.filtered.norm, design, plot = FALSE)
fit <- lmFit(v.DEGList.filtered.norm, design)
contrast.matrix <- makeContrasts(infection_with_WT = crypto.wt - uninfected,
                                 infection_with_Mut = crypto.mut - uninfected,
                                 levels=design)

fits <- contrasts.fit(fit, contrast.matrix)
ebFit <- eBayes(fits)
myTopHits <- topTable(ebFit, adjust ="BH", coef=1, number=40000, sort.by="logFC")
myTopHits <- as_tibble(myTopHits, rownames = "geneSymbol")
ggplot(myTopHits, aes(y=-log10(adj.P.Val), x=logFC, text = paste("Symbol:", geneSymbol))) +
  geom_point(size=2) +
  ylim(-0.5,12) +
  geom_hline(yintercept = -log10(0.01), linetype="longdash", colour="grey", size=1) +
  geom_vline(xintercept = 1, linetype="longdash", colour="#BE684D", size=1) +
  geom_vline(xintercept = -1, linetype="longdash", colour="#2C467A", size=1) +
  labs(title="Volcano plot",
       subtitle = "C. parvum infected vs. naive (HCT-8 cells)",
       caption=paste0("produced on ", Sys.time())) +
  theme_bw()

```


Column {.tabset data-width=500}
------------------------------------------------------------------------------

### DEG table

```{r DEG table}
results <- decideTests(ebFit, method="global", adjust.method="BH", p.value=0.05, lfc=1)
colnames(v.DEGList.filtered.norm$E) <- sampleLabels
diffGenes <- v.DEGList.filtered.norm$E[results[,1] !=0 | results[,2] !=0,]
diffGenes.df <- as_tibble(diffGenes, rownames = "geneSymbol")
datatable(diffGenes.df, 
          extensions = c('KeyTable', "FixedHeader"), 
          caption = 'Table 2: DEGs for infected (wt Crypto) vs control',
          options = list(keys = TRUE, searchHighlight = TRUE, pageLength = 10, lengthMenu = c("10", "25", "50", "100"))) %>%
  formatRound(columns=c(1:10), digits=2)

```

### heatmap

```{r heatmap}
library(gplots) #the heatmap2 function in this package is a primary tool for making heatmaps
library(RColorBrewer) #need colors to make heatmaps
library(heatmaply) #for making interactive heatmaps using plotly
myheatcolors2 <- colorRampPalette(colors=c("yellow","white","blue"))(100)
clustRows <- hclust(as.dist(1-cor(t(diffGenes), method="pearson")), method="complete") #cluster rows by pearson correlation
clustColumns <- hclust(as.dist(1-cor(diffGenes, method="spearman")), method="complete")
module.assign <- cutree(clustRows, k=2)
module.color <- rainbow(length(unique(module.assign)), start=0.1, end=0.9) 
module.color <- module.color[as.vector(module.assign)] 
heatmap.2(diffGenes, 
          Rowv=as.dendrogram(clustRows), 
          Colv=as.dendrogram(clustColumns),
          RowSideColors=module.color,
          col=myheatcolors2, scale='row', labRow=NA,
          density.info="none", trace="none",  
          cexRow=1, cexCol=1, margins=c(8,20)) 
```



About {data-orientation=rows data-icon="fa-comment-alt"}
===================================== 

Row {data-height=1000}
------------------------------------------------------------------------------

Use this page to describe the results in your own words.  Some things to think about:

* What are the key takeaways from the analysis?
* What types of analyses would you want to do next?
* Based on your analysis, what wet-lab experiments would you pursue?
* How could you expand on or otherwise enhance this data dashboard?

