---
title: "<center> The transcriptional response during human cutaneous leishmaniasis <center>"
author: "<center> DIYtranscriptomics class<center><br>"
date: "<center> _`r Sys.Date()`_ <center>"
output:
  html_document:
    code_folding: show
    df_print: paged
    theme: yeti
    highlight: tango
    toc: yes
    toc_float:
      collapsed: false
      smooth_scroll: false
    number_sections: true
  pdf_document:
    fig_caption: yes
    toc: yes
---

```{r setup, include=FALSE}
library(rmarkdown)
library(tinytex)
library(knitr)
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

## Introduction

During the Spring 2020 offering of **[DIYtranscriptomics](http://diytranscriptomics.com/)**, we analyzed a subset of patients and healthy controls from [Amorim et al., 2019](https://doi.org/10.1126/scitranslmed.aax4204).  This reproducible and dynamic report was created using Rmarkdown and the [Knitr package](https://yihui.name/knitr/), and summarizes the basic code and outputs (plots, tables, etc) produced during the course. 

***

## R packages used

A variety of R packages was used for this analysis.  All graphics and data wrangling were handled using the [tidyverse suite of packages](https://www.tidyverse.org/).  All packages used are available from the Comprehensive R Archive Network (CRAN), Bioconductor.org, or Github.  

***

## Read mapping

### Aligning raw reads with Kallisto

Raw reads were mapped to the human reference transcriptome using [Kallisto](https://pachterlab.github.io/kallisto/), version 0.46.2.  The quality of raw reads, as well as the results of Kallisto mapping are summarized in [this summary report](http://DIYtranscriptomics.github.io/Data/files/multiqc_report.html) generated using [fastqc](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/) and [multiqc](https://multiqc.info/).


***

### Importing count data into R

After read mapping with Kallisto, [TxImport](https://bioconductor.org/packages/release/bioc/html/tximport.html) was used to read kallisto outputs into the R environment.  Annotation data from Biomart was used to summarize data from transcript-level to gene-level.

```{r step 1 - TxImport}

```

***

## Preprocessing

### Impact of filtering and normalization

```{r step2 - dataWrangling}

```

Filtering was carried out to remove lowly expressed genes. Genes with less than 1 count per million (CPM) in at least 5 or more samples filtered out.  This reduced the number of genes from `r nrow(myDGEList)` to `r nrow(myDGEList.filtered)`.  

***

### table of filtered and normalized data

```{r step 3 - multivariate part 1 (data table)}

```

The table shown below includes expression data for `r nrow(myDGEList.filtered)` genes. You can sort and search the data directly from the table.


***

## PCA plot

```{r step 3 - multivariate part 2 (PCA plot)}

```

***

## Volcano plot

```{r step 5 - diffGenes part 1 (volcano plot)}

```

***

## Table of DEGs

To identify differentially expressed genes, precision weights were first applied to each gene based on its mean-variance relationship using [VOOM](https://genomebiology.biomedcentral.com/articles/10.1186/gb-2014-15-2-r29), then data was normalized using the [TMM method](https://genomebiology.biomedcentral.com/articles/10.1186/gb-2010-11-3-r25) in [EdgeR](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2796818/).  Linear modeling and bayesian stats were employed via [Limma](https://academic.oup.com/nar/article/43/7/e47/2414268) to find genes that were up- or down-regulated in leishmania patients by 4-fold or more, with a false-discovery rate (FDR) of 0.01. 

```{r step 5 - diffGenes part 2 (DEG table)}

```

***

## Heatmaps and modules

Pearson correlation was used to cluster **`r nrow(diffGenes)`** differentially expressed genes, which were then represented as heatmap with the data scaled by Zscore for each row.

```{r step 6 - modules part 1 (heatmap)}

```


```{r step 6 - modules part 2 (upregulated genes)}

```


```{r step 6 - modules part 3 (downregulated genes)}

```



## GO enrichment

GO enrichment for the `r nrow(myTopHits)` genes induced by infection

```{r step 7 - functionalEnrichment part 1 (gostplot for upregulated genes)}


```

```{r step 7 - functionalEnrichment part 2 (gostplot for downregulated genes)}

```

## GSEA

```{r step 7 - functionalEnrichment part 3 (GSEA table)}

```

```{r step 7 - functionalEnrichment part 4 (enrich plot)}

```


```{r step 7 - functionalEnrichment part 5 (bubble plot)}

```

***

## Conclusions

Describe the results in your own words.  Some things to think about:

* What are the key takeaways from the analysis?
* What types of analyses would you want to do next?
* Based on your analysis, are there any wet-lab experiments would might priortize?
* How could you expand on or otherwise enhance this Rmarkdown report?

## Session info

The output from running 'sessionInfo' is shown below and details all packages and version necessary to reproduce the results in this report.

```{r session info}
sessionInfo()
```

